% Chapter 6

\chapter{Related work} 
\label{chap:related} 


\epigraph{\textit{“If you know the enemy and know yourself you need 
                             not fear the results of a hundred battles."}}{Sun Tzu}

\minitoc

\lhead{Chapter 6. \emph{Related Work}} % This is for the header on each page - perhaps a shortened title

%----------------------------------------------------------------------------------------

		This Chapter discusses the works related with this thesis. Section \ref{sec:relhyper} places the industrial
	case study discussed in Chapter \ref{chap:hyper} in the context of existing literature regarding behavioural
	specification, and architectural reconfigurations. Section \ref{sec:relmefresa} discusses other undertakings concerning the formalization
	of component models using proof assistants and other more lightweight specification platforms. Further,
	we also refer material regarding proof systems for reasoning on distributed systems.
	For last, Section \ref{sec:relpainless} compares \textsc{Painless} and its integration with the ProActive middleware
	with other \ac{ADL}s and related tools.
	

\section{On the HyperManager use case}
\label{sec:relhyper}

	
	
	The use of automata-based formalisms to model software systems is a rather common
	approach in the realm of formal methods. This stems from their intuitive usage,
	and tool support. For instance, the maturity attained by the CADP toolbox \cite{garavel:inria-00583776}
		made it a reference tool among the formal methods community.  Several case studies have been published, 
		namely industrial ones addressing other goals than verification. For instance, in \cite{conf/cav/CosteHLS09}
	Coste et. al. discuss performance evaluation for systems and networks on chips. 
	
	In \cite{BHHM:FACS11}, Ameur-Boulifa et. al. discussed the specification and verification
	of a fault-tolerant \ac{GCM} application. Indeed, their approach is akin to the one
	we used for \textsc{The HyperManager} application. Yet, structural reconfigurations
	are not contemplated. In fact, our case study described in Chapter \ref{chap:hyper}
	is the first work addressing the specification and verification of a 
	reconfigurable \ac{GCM} application.
	

		 	More closely related with the formalization of dynamic architecture topologies aspects,		 	
		 we must refer the experiments presented in 
	 \cite{conf/dais/CornejoGMP01}. A dynamic reconfiguration protocol is specified and model-checked,
	 however their focus is on the reconfiguration protocol itself rather than reconfigurable applications.
		Indeed, many works can be found in the literature embracing a behavioural semantics
	approach for the specification and verification of distributed systems. Yet, 
	literature addressing the aspects of reconfigurable applications remains 
	scarce. 
		
 	Nevertheless, we must cite the work around
	    %BIP
	    BIP (Behaviour, Interaction, Priority) \cite{BBB+11a} --- a framework 
	 encompassing rigorous design principles. It allows the description of the 
	 coordination between components
	 in a layered way. Moreover, it has the particularity of also permitting the 
	 generation of code from its models. Yet, structural reconfigurations are not supported. 
	 Further, as demonstrated in \cite{BBN+10}, their compositional rationale	 allows to 	 
	 combine interactive proof strategies with model-checking.	 
	 	 	
	 	
	%CHARMY	
	Another rather different approach that we must refer is the one followed by tools 
	specifically tailored for architectural specifications. For instance, in \cite{p.2005-1} Inverardi et. al. 
	discusses \textsf{CHARMY}, a framework for designing and validating architectural 
	specifications. It offers a full featured graphical interface with the goal of being more 
	\textit{user friendly} in an industrial context. Still,
	architectural specifications remain of static nature. 
	
		
	
		Looking at the interactive theorem proving arena we can also find some
	related material. In %coq stuff
	\cite{Boyer:2013:RRC:2486788.2486791} Boyer et. al. propose a reconfiguration protocol
	and prove its correctness in the Coq Proof Assistant \cite{09thecoq}. This work however, also focuses on the protocol itself, 
	and not in the behaviour of a reconfigurable application. 
   





\section{On the Mefresa framework}
\label{sec:relmefresa}


	Many approaches regarding the formalization of component models can be found in the literature. 
	In \cite{HKK:FMCO09}, Henrio et al.  formalize a framework for reasoning 
	on the structural aspects of \ac{GCM} component composition mechanized in 
	Isabelle/HOL \cite{Nipkow-Paulson-Wenzel:2002}. Further, they provide a semantics
	for component communication, and illustrate an example concerning request delegation.	 
	Regarding component composition, our approach is considerably different
	in that we provide a semantics for an \textit{operation} language for the building 
	and reconfiguration of \ac{GCM} architectures. Moreover, we define a generic
	datatype \textsf{Net}, that can be used for both communication and behavioural
	aspects.	
	
	
	%CREOL
	Another work involving the use of a proof assistant is the work by Johnsen et. al. 
	on the \textsf{Creol} framework \cite{johnsen06tcs}. \textsf{Creol} focuses on
	the reasoning of distributed systems by providing a high-level object oriented
	modelling language. Its operational semantics are defined in the rewriting logic tool 
	Maude \cite{Maude2:03}. This work however does not contemplate architectural 
	reconfigurations and follows a methodology in the style of a Hoare Logic. 	
		
		
	%TLA proof		
			
	Furthermore, the work around the TLA+ Proof System (TLAPS) \cite{DBLP:journals/corr/abs-1208-5933} 
	is also worth mentioning. TLA+ is a specification language based on set theory and temporal logic,
	with recently added constructs for writing proofs. One of its main strengths is its integrated
	environment allowing to reason about TLA+ specifications using both a model-checker (TLC model-checker),
	and TLAPS. Moreover, it should be noted that TLAPS provides several back-end provers: the proof
	assistant Isabelle \cite{Nipkow-Paulson-Wenzel:2002}, 
	and automated provers such as Zenon \cite{DBLP:conf/lpar/BonichonDD07}.
	However, at this stage of development it does not perform temporal reasoning. Perhaps
	a trace-based approach as the one we illustrated in Chapter \ref{chap:behaviour} would
	suit their needs?
		
	The approach embraced by \textsf{BSP-WHY} \cite{Fortin:2010:BIL:1863482.1863491} also permits 
	the use of several back-end provers. It is tailored for 
	\textsf{BSP} programs \cite{Valiant:1990:BMP:79173.79181}, and relies on the 
	\textsf{Why} tool \cite{Filliatre:2007:WPD:1770351.1770379} to produce 
	the relevant proof obligations w.r.t. to the properties of interest.
	
				
     	
	%most stuff is "just" model checking
	%invited speaker conf. http://www.di.univaq.it/inverard/
	%http://www.di.univaq.it/charmy/
	Indeed, while the use of deductive methods, let it be by means of proof assistants 
	or automated provers, is gaining notoriety, model-checking is still
	the \textit{de facto} formal approach for both industrial and academic undertakings.
	Usually, some form of state machine model is used for the design of the intended
	system,  and then a carefully chosen model-checker as back-end is employed as
	a decision procedure. This is for instance the methodology followed in our Vercors
	platform \cite{Barros20073}. 
	

	%formal spec alloy  -> they recommend Coq
   	A formal specification of the Fractal Component Model has 
   	been proposed in the  Alloy specification language \cite{MERLE:2008:INRIA-00338987:1}. 
	This work proves the consistency of a (set-theoretic) model of Fractal applications. 	 
	Their goal was to clarify the inherent ambiguities of the informal specification presented
  	in \cite{fractalSpec}. Their specification however, is constrained by the first-order relational logic nature 
   	of the Alloy Analyzer. In fact, they point to the use of the Coq Proof Assistant in order 
   	to overcome this limitation.	
	
	
	%era PC ano passado
	For last, let us also mention previous work by Aldrich on the use of a type system at the source code level to 
	enforce architectural constraints \cite{Abi-antoun08staticconformance}. This is a rather different
	approach than ours as it is a direct extension of the Java programming language with type annotations
	that are statically checked. Allowing for flexible and dynamic designs is within the goals of the project,	
	but structural reconfigurations are not supported.


	%\paragraph{}
	%see JNO 

	%in the spirit of dynamic ADLs
	%\paragraph{}
	%may want to refer dynamic ADLs ?


	
	%\paragraph{}
	%Extension to previous work \cite{gaspar:hal-00725291}


\section{On Painless}
\label{sec:relpainless}

	Several proposals for \ac{ADL}s can be found in the literature, each with their own 
	particularities and attempting to 
	address their specific application domain concerns. 
	
	%http://www-systems.cs.st-andrews.ac.uk/wiki/ArchWare/
	The work around the ArchWare \ac{ADL} \cite{DBLP:journals/corr/abs-1006-4829} 
	is of particular interest. They claim that 
	\textit{"software that cannot change is condemned to atrophy"}	
	and introduce the concept of an \textit{active software architecture} in order
	to address this challenge. Based on the higher-order $\pi$-calculus, it provides
	constructs for specifying control flow, communication and dynamic topology.
	Unlike \textsc{Painless}, its syntax exhibits an imperative style 
	and type inference is not supported, thus not promoting concise specifications. 
	
	Nevertheless, it is sufficiently rich to provide executable specifications of active software
	architectures. Moreover, user-defined constraints are supported through the ArchWare 
	Architecture Analysis Language. Yet, their focus is more aimed at the specification
	and analysis of the \ac{ADL}, rather than actual application execution and deployment.
	In our work, the user solely defines the architecture of its application, structural constraints
	are implicit: they are within the mechanized \ac{GCM} specification.	Further, our tool support
	is tightly coupled with the ProActive middleware.
	
		
	%ref Archery
		Also from the realm of process algebras, Archery \cite{conf/facs2/SanchezBR11} is a modelling
	language for software architectural patterns. It is composed by a core language and two extensions:
	\textsf{Archery-Core}, \textsf{Archery-Script} and \textsf{Archery-Structural-Constraint}.
	These permit the specification of structural and behavioural dimensions of architectures,
	the definition of scripts for  reconfiguration, and the formulation of structural constraints,
	respectively. Moreover, a bigraphical semantics is defined for Archery specifications. This grants
	the reduction of the constraint satisfaction verification to a type-checking problem. However,
	this process is not guarantee to be automatic, and type-checking decidability remains as future work.
	
	
	%%check refs....
	%falta darwin..
				
		
	%logic based. http://ce-resd.facom.ufms.br/sbrc/1994/p10.pdf
	As expected, logic is also a natural formalism of choice. For instance, Gerel \cite{endler1992}
	is a generic reconfiguration language including powerful query constructs based on first-order logic.
	Further, its reconfiguration procedures may contain preconditions \textit{\`a la} 
	Hoare Logic \cite{Hoare69anaxiomatic}. These are evaluated by brute force. It is unclear
	how they cope with the inherent undecidability of such task. 
	%paper runtime verif. http://hal.archives-ouvertes.fr/docs/00/64/23/45/PDF/facs2011_preproceedings_9.pdf
	%favs 2011
	%temporal logic to FScript ... undecidable problem in conclusion...


		%graph based:
	Furthermore, the use of a graph-based formalism for the specification of software architectures
	is also rather popular. For instance, in \cite{Bruni:2008:GDA:1424922.1424928} two graph-based
	approaches and related tool support are compared. Another interesting aspect of this work
	regards the use of the Alloy Analyzer \cite{Jackson:2002:ALO:505145.505149}, 
	and Maude \cite{Clavel:2002:MSP:633559.633563} as a means to show the approaches' feasibility. 
		

		More recently, 	Di Cosmo et. al. defined the Aeolus component model \cite{conf/sefm/CosmoZZ12}. Their focus
	is on the automation of cloud-based applications deployment scenarios. Nevertheless, their 
    proposal is still loosely inspired by the Fractal component model \cite{fractalSpec} whose
	most peculiar characteristics are its hierarchical composition nature and reconfiguration capabilities.
	However, while both approaches permit architectural reconfigurations at runtime, 
    its specification is not supported by their \ac{ADL}, it solely contemplates deployment related
    aspects. Moreover, support for
	parametrized specifications is also not covered, forcing the software architect to explicitly 	
	define the application's structure.    
	
		Regarding Fractal,  it is also worth noticing that they try to overcome the lack of
	support for reconfiguration specification	 through 
	Fscript \cite{DAVID:2008:HAL-00468474:1}. Fscript embeds FPath --- a DSL for navigation and querying of
	Fractal architectures --- and acts as a scripting language for reconfiguration strategies. These
	are not evaluated for their validity. Nevertheless, system consistency is ensured by the use of
	\textit{transactions}: a violating reconfiguration is \textit{rolled back}.
	Furthermore, an interesting feature is the support for application-specific architectural
	invariants.


	%ref survey 2/5
	%	scalability / types.. logic etc
	%question "Given system x, what happens when change y occurs?"
	%para tese devo colocar me em todas as tabelas referidas..
		To conclude, let us mention a survey regarding the approaches for the specification of 
	dynamic software architectures \cite{Bradbury:2004:SSD:1075405.1075411}. These are 
	classified into four categories: graph-based, process algebra-based, 
	logic-based, and other. To this end, we place ourselves
	in the \textit{other} category. Indeed, while formal logic is at the foundation of 
	our approach, all the machinery involving requirement/constraint checking is 
	implicit and done automatically by our Coq development. The software architect 
	solely needs to use a language with a declarative trait. %Another important factor referred by 
	%this survey concerns scalability. In general, a decentralized or distributed approach
	%is more likely to scale. This is the case for \textsc{Painless}, as	
	
